---
title: "Blackberry Budbreak GWAS"
subtitle: "Two-Year Tetraploid Pipeline — Y1 (2024–25) & Combined (Y1+Y2)"
author: "Rhys Brock"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "▶ Show code"
    self-contained: true
    theme: cosmo
    fig-width: 13
    fig-height: 7
execute:
  warning: false
  message: false
  cache: false
---

```{=html}
<style>
body { font-family: "Lato", "Helvetica Neue", sans-serif; color: #2c3e50; line-height: 1.7; }
h1, h2, h3, h4 { color: #1a3a4a; font-weight: 700; }
h1 { border-bottom: 3px solid #2e7d52; padding-bottom: 0.3em; margin-top: 1.2em; }
h2 { border-left: 5px solid #2e7d52; padding-left: 0.7em; margin-top: 1.8em; }
h3 { color: #2e7d52; margin-top: 1.3em; }
h4 { color: #1a5fa8; font-size: 1em; }
.banner { background: linear-gradient(135deg, #1a3a4a 0%, #2e7d52 100%); color: white; padding: 1.1em 1.6em; border-radius: 8px; margin: 2em 0 1em; font-size: 1.1em; font-weight: 700; letter-spacing: 0.03em; }
.banner-y1   { background: linear-gradient(135deg, #1a5fa8 0%, #2471b5 100%); color: white; padding: 0.9em 1.4em; border-radius: 8px; margin: 1.5em 0 0.8em; font-size: 1.05em; font-weight: 700; }
.banner-comb { background: linear-gradient(135deg, #c0392b 0%, #e74c3c 100%); color: white; padding: 0.9em 1.4em; border-radius: 8px; margin: 1.5em 0 0.8em; font-size: 1.05em; font-weight: 700; }
.banner-rp   { background: linear-gradient(135deg, #1a6b3a 0%, #27ae60 100%); color: white; padding: 0.9em 1.4em; border-radius: 8px; margin: 1.5em 0 0.8em; font-size: 1.05em; font-weight: 700; }
.banner-ord  { background: linear-gradient(135deg, #b7410e 0%, #e67e22 100%); color: white; padding: 0.9em 1.4em; border-radius: 8px; margin: 1.5em 0 0.8em; font-size: 1.05em; font-weight: 700; }
.info-box  { background:#eaf4ee; border-left:5px solid #2e7d52; border-radius:0 6px 6px 0; padding:0.85em 1.1em; margin:0.9em 0; color:#1a3a4a; }
.warn-box  { background:#fef9ec; border-left:5px solid #e6a817; border-radius:0 6px 6px 0; padding:0.85em 1.1em; margin:0.9em 0; color:#4a3800; }
.stat-box  { background:#eaf1fb; border-left:5px solid #1a5fa8; border-radius:0 6px 6px 0; padding:0.85em 1.1em; margin:0.9em 0; color:#0d2d54; }
table { border-collapse:collapse; width:100%; margin:1em 0; font-size:0.9em; }
thead tr { background:#1a3a4a; color:white; }
thead th  { padding:0.55em 0.9em; text-align:left; }
tbody tr:nth-child(even) { background:#f0f5f2; }
tbody tr:hover { background:#d8ede2; }
tbody td  { padding:0.45em 0.9em; border-bottom:1px solid #cde0d4; }
.badge { display:inline-block; padding:2px 10px; border-radius:12px; font-size:0.78em; font-weight:700; vertical-align:middle; margin-left:6px; }
.badge-y1   { background:#dbeafe; color:#1e40af; border:1px solid #93c5fd; }
.badge-comb { background:#fee2e2; color:#991b1b; border:1px solid #fca5a5; }
.badge-ff   { background:#d1fae5; color:#065f46; border:1px solid #6ee7b7; }
.badge-pf   { background:#fef3c7; color:#92400e; border:1px solid #fcd34d; }
.badge-ffpf { background:#e0e7ff; color:#3730a3; border:1px solid #a5b4fc; }
.badge-rp   { background:#d1fae5; color:#065f46; border:1px solid #34d399; }
.badge-ord  { background:#fde8d8; color:#7d2d00; border:1px solid #f0a070; }
.callout-note .callout-header { background:#d4edda !important; color:#1a3a4a !important; font-weight:600; }
.callout-note { border-left-color:#2e7d52 !important; }
#TOC { border-right:2px solid #2e7d52; padding-right:1em; }
#TOC a { color:#1a3a4a; }
#TOC a:hover { color:#2e7d52; text-decoration:underline; }
</style>
```

---

<div class="banner">📋 Study Overview & Statistical Framework</div>

# Study Overview

## What This Document Does

<div class="info-box">
This pipeline performs GWAS on blackberry budbreak phenotypes using `GWASpoly` for
tetraploid populations. **72 analyses** across 6 blocks (A–F) and three phenotype
encoding sets. Y1 data are analysed directly (blocks A–D); Y2 data contribute
through the combined-year BLUE estimation only and are not run as a standalone year.
Blocks A–D (runs 1–32) are already complete.
</div>

## Run Block Summary

**Runs 1–48 — Original 8 traits per block:**

| Block | Year | Cohort | Covariate | Purpose | Runs |
|-------|------|--------|-----------|---------|------|
| A | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ffpf">FFPF</span> | PF | PF-independent QTL, Y1 | 1–8 |
| B | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ff">FF only</span> | None | FF-background QTL, Y1 | 9–16 |
| C | <span class="badge badge-y1">Y1</span> | <span class="badge badge-pf">PF only</span> | None | PF-background modifiers, Y1 | 17–24 |
| D | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ffpf">FFPF</span> | None | PF locus test, Y1 | 25–32 |
| E | <span class="badge badge-comb">Combined</span> | <span class="badge badge-ffpf">FFPF</span> | PF | Year-stable PF-independent QTL | 33–40 |
| F | <span class="badge badge-comb">Combined</span> | <span class="badge badge-ff">FF only</span> | None | Year-stable FF QTL (highest priority) | 41–48 |

**Runs 49–60 — REP_PROP (2 traits per block):**

| Block | Year | Cohort | Covariate | Runs |
|-------|------|--------|-----------|------|
| A_RP | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ffpf">FFPF</span> | PF | 49–50 |
| B_RP | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ff">FF</span> | None | 51–52 |
| C_RP | <span class="badge badge-y1">Y1</span> | <span class="badge badge-pf">PF</span> | None | 53–54 |
| D_RP | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ffpf">FFPF</span> | None | 55–56 |
| E_RP | <span class="badge badge-comb">Combined</span> | <span class="badge badge-ffpf">FFPF</span> | PF | 57–58 |
| F_RP | <span class="badge badge-comb">Combined</span> | <span class="badge badge-ff">FF</span> | None | 59–60 |

**Runs 61–72 — ORD (2 traits per block):**

| Block | Year | Cohort | Covariate | Runs |
|-------|------|--------|-----------|------|
| A_ORD | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ffpf">FFPF</span> | PF | 61–62 |
| B_ORD | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ff">FF</span> | None | 63–64 |
| C_ORD | <span class="badge badge-y1">Y1</span> | <span class="badge badge-pf">PF</span> | None | 65–66 |
| D_ORD | <span class="badge badge-y1">Y1</span> | <span class="badge badge-ffpf">FFPF</span> | None | 67–68 |
| E_ORD | <span class="badge badge-comb">Combined</span> | <span class="badge badge-ffpf">FFPF</span> | PF | 69–70 |
| F_ORD | <span class="badge badge-comb">Combined</span> | <span class="badge badge-ff">FF</span> | None | 71–72 |

**Trait definitions:**

| Trait | Type | Notes |
|-------|------|-------|
| `BIN_T1` | Binary | 1 = any budbreak at T1, 0 = none |
| `BIN_T2` | Binary | 1 = any budbreak at T2, 0 = none |
| `PERCENT_T1` | Continuous | % buds broken at T1, per-genotype mean |
| `PERCENT_T2` | Continuous | % buds broken at T2, per-genotype mean |
| `BLUE_PERCENT` | Adjusted | Block+Timepoint(+Year) fixed-effect BLUE |
| `COMP_PERCENT` | Composite | Mean of PERCENT_T1 and PERCENT_T2 |
| `PC1` | PCA score | Major axis of T1/T2 covariation |
| `PC2` | PCA score | Contrast axis (early vs late budbreak) |
| `REP_PROP_T1` | Proportion | Fraction of reps showing budbreak at T1 |
| `REP_PROP_T2` | Proportion | Fraction of reps showing budbreak at T2 |
| `ORD_T1` | Ordinal | 0 = 0 reps broken; 1 = 1–2 reps; 2 = 3–4 reps at T1 |
| `ORD_T2` | Ordinal | 0 = 0 reps broken; 1 = 1–2 reps; 2 = 3–4 reps at T2 |

---

<div class="banner">⚙️ Setup</div>

# Setup

## Packages

```{r packages}
pkgs <- c("tidyverse", "data.table", "GWASpoly", "stringr", "lme4", "parallel")
to_install <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)
invisible(lapply(pkgs, library, character.only = TRUE))
```

## File Paths

```{r paths}
y1_pheno_file <- "C:/Users/RhysB/OneDrive/Desktop/CH2 GWAS/Y1_Data_GWAS/Y1_GWAS_Budbreak_Data_Sorted_ForR.csv"
y2_pheno_file <- "C:/Users/RhysB/OneDrive/Desktop/CH2 GWAS/Y2_Data_GWAS/Y2_GWAS_Budbreak_Data_Sorted_ForR.csv"
geno_file     <- "C:/Users/RhysB/OneDrive/Desktop/CH2 GWAS/captureseq2025_GATK_allsamples_postmean_updog_filtered (1).csv"
out_dir       <- "gwaspoly_runsets"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
mp_outdir     <- "C:/Users/RhysB/OneDrive/Desktop/CH2 GWAS/GWASCHILL_OUTPUTS"
dir.create(mp_outdir, showWarnings = FALSE, recursive = TRUE)
cat("[Y1 Pheno ]", y1_pheno_file,
    "\n[Y2 Pheno ]", y2_pheno_file, "(used for combined-year BLUEs only)",
    "\n[Geno     ]", geno_file,
    "\n[Run tables]", normalizePath(out_dir, mustWork = FALSE),
    "\n[PNG output]", normalizePath(mp_outdir, mustWork = FALSE), "\n")
```

---

<div class="banner">🔧 Helper Functions</div>

# Helper Functions

```{r helpers}
.read_csv <- function(path) {
    data.table::fread(path, data.table = FALSE,
        na.strings = c("", "NA", "na", "no cuttings", "No cuttings", "N/A"))
}
get_name_y1 <- function(x) {
    x <- stringr::str_trim(x)
    x <- sub("_\\d+$", "", x)
    x <- sub("_\\d+_\\d+_[A-Da-d]$", "", x)
    x <- sub("_\\d+_[A-Da-d]$", "", x)
    x <- sub("_[A-Da-d]$", "", x)
    x <- sub("PF$", "", x)
    x
}
get_block_y1 <- function(x) toupper(stringr::str_extract(x, "[A-Da-d](?=_\\d+$)"))
get_name_y2  <- function(x) {
    x <- stringr::str_trim(x)
    x <- sub("^(ANG|NOV|WSH|BEN|BSH)_+", "", x)
    x <- sub("_+L\\d+[A-Da-d]_+P[\\d.]+$", "", x)
    x <- sub("\\s*\\([^)]*\\)", "", x)
    x <- stringr::str_trim(x)
    x <- sub("[_ ]+$", "", x)
    x
}
get_block_y2 <- function(x) toupper(stringr::str_extract(x, "(?<=L\\d)[A-Da-d](?=_+P)"))
.clamp <- function(v, maxv) {
    v <- suppressWarnings(as.numeric(v))
    maxv <- suppressWarnings(as.numeric(maxv))
    pmin(pmax(v, 0, na.rm = FALSE), maxv, na.rm = FALSE)
}
save_png_hq <- function(p, filename, w = 14, h = 8, dpi = 600) {
    ggplot2::ggsave(filename, p, width = w, height = h, dpi = dpi,
        units = "in", type = "cairo-png")
}
```

---

<div class="banner">🌱 Phenotype Construction</div>

# Phenotype Construction

## Year 1 Phenotypes (Dec-27-2024 & Jan-8-2025)

```{r y1-pheno}
y1_raw <- .read_csv(y1_pheno_file)
names(y1_raw) <- make.unique(trimws(names(y1_raw)))

y1_t1 <- y1_raw %>%
    dplyr::select(Lug = 1, TotalBuds = 3, Cnt = 4) %>%
    dplyr::filter(!is.na(Lug), Lug != "") %>%
    dplyr::mutate(NAME = get_name_y1(Lug), Block = get_block_y1(Lug),
                  TotalBuds = suppressWarnings(as.numeric(TotalBuds)),
                  Cnt = .clamp(Cnt, TotalBuds)) %>%
    dplyr::filter(!is.na(TotalBuds), TotalBuds > 0) %>%
    dplyr::select(NAME, Block, TotalBuds_T1 = TotalBuds, Cnt_T1 = Cnt)

y1_t2 <- y1_raw %>%
    dplyr::select(Lug = 6, TotalBuds = 8, Cnt = 9) %>%
    dplyr::filter(!is.na(Lug), Lug != "") %>%
    dplyr::mutate(NAME = get_name_y1(Lug),
                  TotalBuds = suppressWarnings(as.numeric(TotalBuds)),
                  Cnt = .clamp(Cnt, TotalBuds)) %>%
    dplyr::filter(!is.na(TotalBuds), TotalBuds > 0) %>%
    dplyr::select(NAME, TotalBuds_T2 = TotalBuds, Cnt_T2 = Cnt)

y1_long <- dplyr::full_join(y1_t1, y1_t2, by = "NAME") %>%
    dplyr::mutate(PERCENT_T1 = 100 * Cnt_T1 / TotalBuds_T1,
                  PERCENT_T2 = 100 * Cnt_T2 / TotalBuds_T2,
                  BIN_T1 = as.integer(Cnt_T1 > 0),
                  BIN_T2 = as.integer(Cnt_T2 > 0),
                  Year = "Y1")
cat("[Y1] Genotypes:", n_distinct(y1_long$NAME),
    "| T1 obs:", sum(!is.na(y1_long$PERCENT_T1)),
    "| T2 obs:", sum(!is.na(y1_long$PERCENT_T2)), "\n")
```

## Year 2 Phenotypes (Jan-1-2026 & Jan-30-2026)

<div class="info-box">
Y2 data are loaded here solely to contribute to the combined-year BLUE estimation.
No Y2-specific GWAS blocks are run — Y2 genotypes have not yet been sequenced.
</div>

```{r y2-pheno}
y2_raw <- .read_csv(y2_pheno_file)
names(y2_raw) <- make.unique(trimws(names(y2_raw)))

y2_t1 <- y2_raw %>%
    dplyr::select(Tag = 1, TotalBuds = 2, Cnt = 4) %>%
    dplyr::filter(!is.na(Tag), Tag != "") %>%
    dplyr::mutate(NAME = get_name_y2(Tag), Block = get_block_y2(Tag),
                  TotalBuds = suppressWarnings(as.numeric(TotalBuds)),
                  Cnt = .clamp(Cnt, TotalBuds)) %>%
    dplyr::filter(!is.na(TotalBuds), TotalBuds > 0) %>%
    dplyr::select(NAME, Block, TotalBuds_T1 = TotalBuds, Cnt_T1 = Cnt)

y2_t2 <- y2_raw %>%
    dplyr::select(Tag = 6, TotalBuds = 7, Cnt = 9) %>%
    dplyr::filter(!is.na(Tag), Tag != "") %>%
    dplyr::mutate(NAME = get_name_y2(Tag),
                  TotalBuds = suppressWarnings(as.numeric(TotalBuds)),
                  Cnt = .clamp(Cnt, TotalBuds)) %>%
    dplyr::filter(!is.na(TotalBuds), TotalBuds > 0) %>%
    dplyr::select(NAME, TotalBuds_T2 = TotalBuds, Cnt_T2 = Cnt)

y2_long <- dplyr::full_join(y2_t1, y2_t2, by = "NAME") %>%
    dplyr::mutate(PERCENT_T1 = 100 * Cnt_T1 / TotalBuds_T1,
                  PERCENT_T2 = 100 * Cnt_T2 / TotalBuds_T2,
                  BIN_T1 = as.integer(Cnt_T1 > 0),
                  BIN_T2 = as.integer(Cnt_T2 > 0),
                  Year = "Y2")
cat("[Y2] Genotypes:", n_distinct(y2_long$NAME),
    "(used for combined-year BLUEs only — no Y2 GWAS blocks)\n")
```

## PF Status Assignment

```{r pf-assignment}
all_names <- union(unique(y1_long$NAME), unique(y2_long$NAME))
PF_tbl <- tibble::tibble(NAME = all_names) %>%
    dplyr::mutate(PF = dplyr::case_when(
        startsWith(NAME, "APF") ~ 1L,
        startsWith(NAME, "A-")  ~ 0L,
        NAME == "NC430"         ~ 0L,
        NAME %in% c("PrimeArkFreedom","PrimeArkHorizon","Superlicious",
                    "APF-153T","APF-268") ~ 1L,
        NAME %in% c("Osage","Ouachita","Natchez","Apache","Ponca","Caddo",
                    "Navaho","Lochness","Immaculate","Von","Tupy","Stella",
                    "Thunderhead","Celestial","Twilight","A-2362T","A-1905T",
                    "A-2241T","A-1798T","A-2538T","A-2428T","A-1172T",
                    "A-2312","A-2718T") ~ 0L,
        TRUE ~ NA_integer_))
cat("[PF] PF=1:", sum(PF_tbl$PF==1,na.rm=TRUE),
    "| PF=0:", sum(PF_tbl$PF==0,na.rm=TRUE),
    "| PF=NA:", sum(is.na(PF_tbl$PF)), "\n")
unclass <- dplyr::filter(PF_tbl, is.na(PF))$NAME
if (length(unclass) > 0)
    warning("[PF] Unclassified: ", paste(unclass, collapse = ", "))
```

## BLUE Estimation (Per-Year & Combined)

<div class="stat-box">
**BLUE model: `PERCENT ~ Block + Timepoint + NAME`** for Y1. For combined-year:
`PERCENT ~ Year + Block + Timepoint + NAME`. Year is a fixed effect that absorbs
systematic season differences before estimating genotypic values — the resulting
NAME coefficients are the year-adjusted BLUEs passed to GWASpoly.
</div>

```{r blue-models}
compute_blue <- function(df_wide, label) {
    long <- df_wide %>%
        dplyr::select(NAME, Block, PERCENT_T1, PERCENT_T2) %>%
        tidyr::pivot_longer(c(PERCENT_T1, PERCENT_T2), names_to = "Timepoint", values_to = "PERCENT") %>%
        dplyr::filter(!is.na(PERCENT), is.finite(PERCENT)) %>%
        dplyr::mutate(NAME = factor(NAME), Block = factor(Block), Timepoint = factor(Timepoint))
    cat("[BLUE", label, "] obs:", nrow(long), "| genotypes:", n_distinct(long$NAME), "\n")
    mod <- tryCatch(lm(PERCENT ~ Block + Timepoint + NAME, data = long),
                    error = function(e) { message("[BLUE", label, " FALLBACK] ", conditionMessage(e)); NULL })
    if (!is.null(mod)) {
        co <- coef(mod); val <- co[grepl("^NAME", names(co))]
        tibble::tibble(NAME = gsub("^NAME", "", names(val)), BLUE_PERCENT = as.numeric(val))
    } else {
        long %>% dplyr::group_by(NAME) %>%
            dplyr::summarise(BLUE_PERCENT = mean(PERCENT, na.rm = TRUE), .groups = "drop")
    }
}

blue_y1 <- compute_blue(y1_long, "Y1")

comb_long <- dplyr::bind_rows(
    y1_long %>% dplyr::select(NAME, Block, PERCENT_T1, PERCENT_T2, Year),
    y2_long %>% dplyr::select(NAME, Block, PERCENT_T1, PERCENT_T2, Year)
) %>%
    tidyr::pivot_longer(c(PERCENT_T1, PERCENT_T2), names_to = "Timepoint", values_to = "PERCENT") %>%
    dplyr::filter(!is.na(PERCENT), is.finite(PERCENT)) %>%
    dplyr::mutate(NAME = factor(NAME), Block = factor(Block),
                  Year = factor(Year), Timepoint = factor(Timepoint))
cat("[BLUE Combined] obs:", nrow(comb_long), "| genotypes:", n_distinct(comb_long$NAME), "\n")

mod_comb <- tryCatch(lm(PERCENT ~ Year + Block + Timepoint + NAME, data = comb_long),
                     error = function(e) { message("[BLUE Combined FAIL] ", conditionMessage(e)); NULL })
if (!is.null(mod_comb)) {
    co <- coef(mod_comb); val <- co[grepl("^NAME", names(co))]
    blue_comb <- tibble::tibble(NAME = gsub("^NAME", "", names(val)), BLUE_PERCENT = as.numeric(val))
} else {
    blue_comb <- comb_long %>% dplyr::group_by(NAME) %>%
        dplyr::summarise(BLUE_PERCENT = mean(PERCENT, na.rm = TRUE), .groups = "drop")
}
cat("[BLUE Combined] genotypes:", nrow(blue_comb), "\n")
```

## Composite Traits, REP_PROP, ORD & PCA

```{r pca-traits}
# BIN_T1/T2:      mean of per-rep 0/1 binary
# REP_PROP_T1/T2: same, explicitly named as proportion of reps
# ORD_T1/T2:      0 = 0 reps broken, 1 = 1-2 reps broken, 2 = 3-4 reps broken
#                 Added per advisor suggestion (0 vs 1-2 vs 3+ binning)

make_trait_table <- function(df_long, label) {
    means <- df_long %>%
        dplyr::group_by(NAME) %>%
        dplyr::summarise(
            across(c(BIN_T1, BIN_T2, PERCENT_T1, PERCENT_T2), \(x) mean(x, na.rm = TRUE)),
            REP_PROP_T1 = mean(BIN_T1, na.rm = TRUE),
            REP_PROP_T2 = mean(BIN_T2, na.rm = TRUE),
            .groups = "drop") %>%
        dplyr::mutate(
            COMP_PERCENT = rowMeans(cbind(PERCENT_T1, PERCENT_T2), na.rm = TRUE),
            ORD_T1 = dplyr::case_when(REP_PROP_T1 == 0   ~ 0L,
                                       REP_PROP_T1 <= 0.5 ~ 1L, TRUE ~ 2L),
            ORD_T2 = dplyr::case_when(REP_PROP_T2 == 0   ~ 0L,
                                       REP_PROP_T2 <= 0.5 ~ 1L, TRUE ~ 2L))

    pca_in  <- means %>% dplyr::filter(complete.cases(dplyr::select(., PERCENT_T1, PERCENT_T2)))
    pca_mat <- scale(pca_in %>% dplyr::select(PERCENT_T1, PERCENT_T2))
    pca_res <- prcomp(pca_mat, center = FALSE, scale. = FALSE)
    var_exp <- summary(pca_res)$importance["Proportion of Variance", ]
    cat("[PCA", label, "] PC1:", round(var_exp[1]*100,1), "% | PC2:", round(var_exp[2]*100,1), "%\n")
    scores  <- tibble::tibble(NAME = pca_in$NAME, PC1 = pca_res$x[,1], PC2 = pca_res$x[,2])
    dplyr::left_join(means, scores, by = "NAME")
}

traits_y1   <- make_trait_table(y1_long, "Y1")
traits_comb <- make_trait_table(dplyr::bind_rows(y1_long, y2_long), "Combined")
```

## Master Phenotype Tables

```{r master-pheno}
make_master <- function(traits_tbl, blue_tbl, pf_tbl) {
    traits_tbl %>%
        dplyr::left_join(blue_tbl %>% dplyr::select(NAME, BLUE_PERCENT), by = "NAME") %>%
        dplyr::left_join(pf_tbl, by = "NAME")
}
PH_Y1   <- make_master(traits_y1,   blue_y1,   PF_tbl)
PH_COMB <- make_master(traits_comb, blue_comb, PF_tbl)
cat("[PH_Y1  ]", nrow(PH_Y1), "genotypes | PF=1:", sum(PH_Y1$PF==1,na.rm=TRUE),
    "| PF=0:", sum(PH_Y1$PF==0,na.rm=TRUE), "\n")
cat("[PH_COMB]", nrow(PH_COMB), "genotypes | PF=1:", sum(PH_COMB$PF==1,na.rm=TRUE),
    "| PF=0:", sum(PH_COMB$PF==0,na.rm=TRUE), "\n")
```

## Cultivar Name Translation

```{r cultivar-aliases}
cultivar_aliases <- c(
    "Osage"="A-2362T", "Ouachita"="A-1905T", "Natchez"="A-2241T",
    "Apache"="A-1798T", "Navaho"="A-1172T", "Caddo"="A-2428T",
    "Ponca"="A-2538T", "Stella"="A-2312", "Immaculate"="A-2718T",
    "PrimeArkFreedom"="APF-153T", "PrimeArkHorizon"="APF-268",
    "Superlicious"="A-2500TN", "A-2868TN"="A-2868TN-2"
)
translate_names <- function(ph, alias_vec) { ph$NAME <- dplyr::recode(ph$NAME, !!!alias_vec); ph }
PH_Y1   <- translate_names(PH_Y1,   cultivar_aliases)
PH_COMB <- translate_names(PH_COMB, cultivar_aliases)
cat("[ALIAS] Mapped", length(cultivar_aliases), "cultivar names.\n")
cat("[ALIAS] NOTE: APF-708T not in marker file — excluded.\n")
```

---

<div class="banner">🧬 Genotype Harmonization</div>

# Genotype Harmonization

```{r geno-harmonize}
cat("[GENO] Reading genotype matrix...\n")
G <- data.table::fread(geno_file)
names(G)[1:3] <- c("marker", "chrom", "pos")
geno_samples  <- setdiff(names(G), c("marker", "chrom", "pos"))
cat("[GENO] Markers:", nrow(G), "| Samples:", length(geno_samples), "\n")

norm_id   <- function(x) toupper(gsub("[-_ ]", "", stringr::str_trim(x)))
all_pheno <- unique(c(PH_Y1$NAME, PH_COMB$NAME))
geno_map  <- tibble::tibble(orig = geno_samples, key = norm_id(geno_samples))
pheno_map <- tibble::tibble(NAME = all_pheno, key = norm_id(all_pheno))
join_map  <- dplyr::inner_join(geno_map, pheno_map, by = "key")
cat("[GENO-HARM] Overlap:", nrow(join_map), "samples\n")
if (nrow(join_map) < 20) stop("FATAL: fewer than 20 overlapping samples.")

for (i in seq_len(nrow(join_map)))
    data.table::setnames(G, old = join_map$orig[i], new = join_map$NAME[i])

final_samples <- intersect(setdiff(names(G), c("marker","chrom","pos")), all_pheno)
cat("[GENO-HARM] Final usable samples:", length(final_samples), "\n")
G_sub    <- G[, c("marker","chrom","pos", final_samples), with = FALSE]
geno_out <- file.path(out_dir, "geno_numeric_tetra.csv")
data.table::fwrite(G_sub, geno_out)
cat("[GENO-HARM] Harmonised genotype file written.\n")
```

## Year 1 Genotype Match Diagnostic

```{r geno-diagnostic}
geno_header <- data.table::fread(geno_file, nrows = 0)
geno_cols   <- names(geno_header)[-(1:3)]
y1_names    <- sort(unique(PH_Y1$NAME))
geno_norm   <- norm_id(geno_cols)
pheno_norm  <- norm_id(y1_names)
cat("[GENO DIAGNOSTIC] Marker file samples:", length(geno_cols), "\n")
cat("[PHENO Y1] Unique names:", length(y1_names), "\n\n")

diag_tbl <- tibble::tibble(Phenotype_NAME = y1_names, Pheno_Norm = pheno_norm) %>%
    dplyr::mutate(
        Exact_Match = Phenotype_NAME %in% geno_cols,
        Fuzzy_Match = Pheno_Norm     %in% geno_norm,
        Geno_Column = dplyr::case_when(
            Exact_Match                ~ Phenotype_NAME,
            Fuzzy_Match & !Exact_Match ~ geno_cols[match(Pheno_Norm, geno_norm)],
            TRUE                       ~ NA_character_),
        Status = dplyr::case_when(
            Exact_Match                ~ "Exact match",
            Fuzzy_Match & !Exact_Match ~ "Fuzzy match (name differs — verify)",
            TRUE                       ~ "NO MATCH — excluded from GWAS"))

no_match_idx <- which(!diag_tbl$Fuzzy_Match)
diag_tbl$Closest_Geno_Name <- NA_character_
if (length(no_match_idx) > 0)
    diag_tbl$Closest_Geno_Name[no_match_idx] <- sapply(
        diag_tbl$Pheno_Norm[no_match_idx],
        function(pn) geno_cols[which.min(utils::adist(pn, geno_norm))])

cat("Match summary:\n"); print(table(diag_tbl$Status)); cat("\n")
no_match_out <- dplyr::filter(diag_tbl, !Fuzzy_Match) %>%
    dplyr::select(Phenotype_NAME, Closest_Geno_Name, Status)
if (nrow(no_match_out) == 0) cat("All Y1 genotypes found in marker file.\n") else {
    cat(nrow(no_match_out), "Y1 genotypes with NO match:\n\n"); print(no_match_out, n = Inf) }

knitr::kable(
    diag_tbl %>% dplyr::select(Phenotype_NAME, Geno_Column, Closest_Geno_Name, Status) %>%
        dplyr::arrange(Status, Phenotype_NAME),
    col.names = c("Y1 Phenotype NAME","Matched Geno Column","Closest Geno Name","Status"),
    caption = "Y1 genotype match status. Y2 excluded (not yet sequenced).")
```

---

<div class="banner">📐 Run Configuration</div>

# Run Configuration

```{r run-config}
TRAITS <- c("BIN_T1","BIN_T2","PERCENT_T1","PERCENT_T2",
            "BLUE_PERCENT","COMP_PERCENT","PC1","PC2")

blocks <- tibble::tribble(
    ~Block, ~Cohort, ~Covariate, ~Dataset,    ~Start,
    "A",    "FFPF",  "PF",       "Y1",          1,
    "B",    "FF",    "None",     "Y1",          9,
    "C",    "PF",    "None",     "Y1",         17,
    "D",    "FFPF",  "None",     "Y1",         25,
    "E",    "FFPF",  "PF",       "Combined",   33,
    "F",    "FF",    "None",     "Combined",   41
)

run_config_orig <- blocks %>%
    dplyr::rowwise() %>%
    dplyr::mutate(Traits = list(TRAITS)) %>%
    tidyr::unnest(Traits) %>%
    dplyr::rename(Trait = Traits) %>%
    dplyr::mutate(
        run_id = Start + match(Trait, TRAITS) - 1,
        Transformation = dplyr::case_when(
            startsWith(Trait, "BIN")     ~ "Binary (any budbreak)",
            startsWith(Trait, "PERCENT") ~ "Percent (0-100), per-genotype mean",
            Trait == "BLUE_PERCENT"      ~ "BLUE: Block+Timepoint(+Year)-adjusted",
            Trait == "COMP_PERCENT"      ~ "Composite: mean of both timepoints",
            Trait %in% c("PC1","PC2")   ~ "Phenotypic PCA score")) %>%
    dplyr::select(run_id, Block, Cohort, Covariate, Dataset, Trait, Transformation)

TRAITS_RP <- c("REP_PROP_T1","REP_PROP_T2")
blocks_rp <- tibble::tribble(
    ~Block,  ~Cohort, ~Covariate, ~Dataset,    ~Start,
    "A_RP",  "FFPF",  "PF",       "Y1",          49,
    "B_RP",  "FF",    "None",     "Y1",          51,
    "C_RP",  "PF",    "None",     "Y1",          53,
    "D_RP",  "FFPF",  "None",     "Y1",          55,
    "E_RP",  "FFPF",  "PF",       "Combined",    57,
    "F_RP",  "FF",    "None",     "Combined",    59
)
run_config_rp <- blocks_rp %>%
    dplyr::rowwise() %>%
    dplyr::mutate(Traits = list(TRAITS_RP)) %>%
    tidyr::unnest(Traits) %>%
    dplyr::rename(Trait = Traits) %>%
    dplyr::mutate(run_id = Start + match(Trait, TRAITS_RP) - 1,
                  Transformation = "Proportion of reps with any budbreak (0.0-1.0)") %>%
    dplyr::select(run_id, Block, Cohort, Covariate, Dataset, Trait, Transformation)

TRAITS_ORD <- c("ORD_T1","ORD_T2")
blocks_ord <- tibble::tribble(
    ~Block,   ~Cohort, ~Covariate, ~Dataset,    ~Start,
    "A_ORD",  "FFPF",  "PF",       "Y1",          61,
    "B_ORD",  "FF",    "None",     "Y1",          63,
    "C_ORD",  "PF",    "None",     "Y1",          65,
    "D_ORD",  "FFPF",  "None",     "Y1",          67,
    "E_ORD",  "FFPF",  "PF",       "Combined",    69,
    "F_ORD",  "FF",    "None",     "Combined",    71
)
run_config_ord <- blocks_ord %>%
    dplyr::rowwise() %>%
    dplyr::mutate(Traits = list(TRAITS_ORD)) %>%
    tidyr::unnest(Traits) %>%
    dplyr::rename(Trait = Traits) %>%
    dplyr::mutate(run_id = Start + match(Trait, TRAITS_ORD) - 1,
                  Transformation = "Ordinal: 0=no reps broken, 1=1-2 reps, 2=3-4 reps") %>%
    dplyr::select(run_id, Block, Cohort, Covariate, Dataset, Trait, Transformation)

run_config <- dplyr::bind_rows(run_config_orig, run_config_rp, run_config_ord)
cat("[CONFIG] Total runs:", nrow(run_config),
    "(1-48: original | 49-60: REP_PROP | 61-72: ORD)\n")
knitr::kable(run_config, caption = "72-run GWAS configuration")

get_cohort_df <- function(dataset, cohort) {
    if (!exists("PH_Y1") || !exists("PH_COMB"))
        stop("Phenotype tables not found. Run all setup chunks first.")
    ph <- switch(dataset, "Y1" = PH_Y1, "Combined" = PH_COMB,
                 stop("Unknown dataset: ", dataset))
    switch(cohort,
        "FFPF" = dplyr::filter(ph, !is.na(PF)),
        "FF"   = dplyr::filter(ph, !is.na(PF), PF == 0),
        "PF"   = dplyr::filter(ph, !is.na(PF), PF == 1),
        stop("Unknown cohort: ", cohort))
}
```

---

<div class="banner">🔬 GWAS Engine</div>

# GWAS Engine

```{r gwas-function}
run_gwas_single <- function(run_id, force = FALSE) {
    cfg <- dplyr::filter(run_config, run_id == !!run_id)
    if (nrow(cfg) == 0) { message("[SKIP] Run ", run_id, " not in config."); return(invisible(NULL)) }
    trait <- cfg$Trait; cov <- cfg$Covariate
    png_path <- file.path(mp_outdir,
        paste0(sprintf("%03d", run_id), "__", cfg$Block, "__", cfg$Dataset,
               "__", cfg$Cohort, "__", trait, ".png"))
    if (file.exists(png_path) && !force) {
        message("[SKIP] Run ", sprintf("%03d", run_id), " PNG exists. force=TRUE to rerun.")
        return(invisible(NULL))
    }
    df <- get_cohort_df(cfg$Dataset, cfg$Cohort)
    if (!trait %in% names(df)) { message("[SKIP] Trait missing. Run ", run_id); return(invisible(NULL)) }
    n_obs <- sum(!is.na(df[[trait]]))
    if (n_obs < 20) { message("[SKIP] Only ", n_obs, " obs. Run ", run_id); return(invisible(NULL)) }
    df_use <- df %>% dplyr::select(NAME, dplyr::all_of(trait)) %>%
        dplyr::filter(!is.na(.data[[trait]]))
    pheno_out <- file.path(out_dir, paste0("pheno_run", sprintf("%03d", run_id), ".csv"))
    data.table::fwrite(df_use, pheno_out)
    data_gp <- GWASpoly::read.GWASpoly(ploidy = 4, pheno.file = pheno_out,
        geno.file = geno_out, format = "numeric", n.traits = 1)
    if (cov == "PF")
        data_gp@fixed <- data.frame(PF = factor(df$PF[match(rownames(data_gp@pheno), df$NAME)]))
    data_gp <- GWASpoly::set.K(data_gp, LOCO = TRUE)
    params   <- if (cov == "PF") GWASpoly::set.params(fixed = "PF", fixed.type = "factor") else GWASpoly::set.params()
    fit <- GWASpoly::GWASpoly(data = data_gp, traits = trait,
        models = c("additive","1-dom","2-dom","general"), params = params,
        n.core = max(1L, parallel::detectCores() - 1L))
    thr <- GWASpoly::set.threshold(fit, method = "M.eff", level = 0.05)
    base <- file.path(out_dir, paste0("run", sprintf("%03d", run_id), "_", cfg$Block, "_", trait))
    GWASpoly::write.GWASpoly(thr, trait = trait, filename = paste0(base, "_scores.csv"), what = "scores")
    qtl <- GWASpoly::get.QTL(thr, traits = trait, models = "additive", bp.window = 5e6)
    data.table::fwrite(qtl, paste0(base, "_QTL.csv"))
    p <- GWASpoly::manhattan.plot(thr, traits = trait) +
        ggplot2::ggtitle(
            label    = paste0("Run ", sprintf("%03d", run_id), " | Block ", cfg$Block,
                              " | ", cfg$Dataset, " | Cohort: ", cfg$Cohort, " | Trait: ", trait),
            subtitle = paste0(cfg$Transformation,
                              if (cov != "None") paste0(" | Covariate: ", cov) else "")) +
        ggplot2::theme(plot.title    = ggplot2::element_text(size = 13, face = "bold"),
                       plot.subtitle = ggplot2::element_text(size = 11, color = "grey40"))
    save_png_hq(p, png_path)
    message("[DONE] Run ", sprintf("%03d", run_id), " | ", cfg$Block, " | ",
            cfg$Dataset, " | ", trait, "\n  ->  ", png_path)
}
```

---

<div class="banner">🚀 Generate Manhattan Plots</div>

# Generate Manhattan Plots

<div class="warn-box">
⚠️ **Run interactively in RStudio — NOT during document render.**
Runs 1–32 are already complete. Use `run_missing_manhattan_plots()` to continue.
</div>

## Run All 72 Plots

```{r generate-all, eval=FALSE}
for (i in run_config$run_id) run_gwas_single(i)
```

## Run Only Missing Plots

```{r run-missing, eval=FALSE}
run_missing_manhattan_plots <- function() {
    status <- run_config %>%
        dplyr::rowwise() %>%
        dplyr::mutate(
            png_path = file.path(mp_outdir,
                paste0(sprintf("%03d", run_id), "__", Block, "__", Dataset,
                       "__", Cohort, "__", Trait, ".png")),
            exists = file.exists(png_path)) %>%
        dplyr::ungroup()
    done    <- dplyr::filter(status, exists)
    missing <- dplyr::filter(status, !exists)
    cat("[STATUS]", nrow(done), "complete |", nrow(missing), "missing\n\n")
    if (nrow(missing) == 0) { message("[OK] All 72 plots exist."); return(invisible(NULL)) }
    print(missing %>% dplyr::select(run_id, Block, Cohort, Dataset, Trait))
    for (i in missing$run_id) { message("\n[RUNNING] Run ", sprintf("%03d", i)); run_gwas_single(i) }
    message("\n[COMPLETE] All missing plots generated.")
}
run_missing_manhattan_plots()
```

## Run a Single Block

```{r run-one-block, eval=FALSE}
run_block <- function(block_letter) {
    ids <- dplyr::filter(run_config, Block == block_letter)$run_id
    for (i in ids) run_gwas_single(i)
}
run_block("A")      # Y1, FFPF + PF covariate (runs 1-8)  — DONE
run_block("E")      # Combined, FFPF + PF covariate (runs 33-40)
run_block("F")      # Combined, FF only (runs 41-48)
run_block("A_ORD")  # Y1, FFPF, ORD (runs 61-62)
run_block("F_ORD")  # Combined, FF only, ORD (runs 71-72)
```

---

<div class="banner">📊 Results — Manhattan Plots</div>

# Results — Manhattan Plots

```{r display-helper, include=FALSE}
show_manhattan <- function(run_id) {
    cfg <- dplyr::filter(run_config, run_id == !!run_id)
    if (nrow(cfg) == 0) { cat("> Run", run_id, "not found.\n\n"); return(invisible()) }
    png_path <- file.path(mp_outdir,
        paste0(sprintf("%03d", run_id), "__", cfg$Block, "__", cfg$Dataset,
               "__", cfg$Cohort, "__", cfg$Trait, ".png"))
    if (file.exists(png_path)) knitr::include_graphics(png_path) else
        cat("> Plot not yet generated. Run `run_gwas_single(", run_id, ")` then re-render.\n\n")
}
```

---

<div class="banner-y1">📅 Year 1 Results (Dec-27-2024 & Jan-8-2025)</div>

## Block A — <span class="badge badge-y1">Y1</span> <span class="badge badge-ffpf">FFPF</span> + PF Covariate (Runs 1–8) {#block-a}

<div class="stat-box">**Purpose:** PF-independent QTL, Y1. Compare with Block D — peaks in both are confirmed PF-independent.</div>

::: {.callout-note collapse="true" title="Run 001 | Y1 | FFPF | BIN_T1 | Cov: PF"}
```{r p001, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(1)
```
:::
::: {.callout-note collapse="true" title="Run 002 | Y1 | FFPF | BIN_T2 | Cov: PF"}
```{r p002, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(2)
```
:::
::: {.callout-note collapse="true" title="Run 003 | Y1 | FFPF | PERCENT_T1 | Cov: PF"}
```{r p003, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(3)
```
:::
::: {.callout-note collapse="true" title="Run 004 | Y1 | FFPF | PERCENT_T2 | Cov: PF"}
```{r p004, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(4)
```
:::
::: {.callout-note collapse="true" title="Run 005 | Y1 | FFPF | BLUE_PERCENT | Cov: PF"}
```{r p005, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(5)
```
:::
::: {.callout-note collapse="true" title="Run 006 | Y1 | FFPF | COMP_PERCENT | Cov: PF"}
```{r p006, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(6)
```
:::
::: {.callout-note collapse="true" title="Run 007 | Y1 | FFPF | PC1 | Cov: PF"}
```{r p007, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(7)
```
:::
::: {.callout-note collapse="true" title="Run 008 | Y1 | FFPF | PC2 | Cov: PF"}
```{r p008, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(8)
```
:::

---

## Block B — <span class="badge badge-y1">Y1</span> <span class="badge badge-ff">FF Only</span> (Runs 9–16) {#block-b}

<div class="stat-box">**Purpose:** FF-background QTL, Y1. PF locus has zero variance in this cohort.</div>

::: {.callout-note collapse="true" title="Run 009 | Y1 | FF | BIN_T1"}
```{r p009, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(9)
```
:::
::: {.callout-note collapse="true" title="Run 010 | Y1 | FF | BIN_T2"}
```{r p010, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(10)
```
:::
::: {.callout-note collapse="true" title="Run 011 | Y1 | FF | PERCENT_T1"}
```{r p011, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(11)
```
:::
::: {.callout-note collapse="true" title="Run 012 | Y1 | FF | PERCENT_T2"}
```{r p012, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(12)
```
:::
::: {.callout-note collapse="true" title="Run 013 | Y1 | FF | BLUE_PERCENT"}
```{r p013, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(13)
```
:::
::: {.callout-note collapse="true" title="Run 014 | Y1 | FF | COMP_PERCENT"}
```{r p014, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(14)
```
:::
::: {.callout-note collapse="true" title="Run 015 | Y1 | FF | PC1"}
```{r p015, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(15)
```
:::
::: {.callout-note collapse="true" title="Run 016 | Y1 | FF | PC2"}
```{r p016, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(16)
```
:::

---

## Block C — <span class="badge badge-y1">Y1</span> <span class="badge badge-pf">PF Only</span> (Runs 17–24) {#block-c}

<div class="stat-box">**Purpose:** PF-background modifier loci, Y1.</div>

::: {.callout-note collapse="true" title="Run 017 | Y1 | PF | BIN_T1"}
```{r p017, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(17)
```
:::
::: {.callout-note collapse="true" title="Run 018 | Y1 | PF | BIN_T2"}
```{r p018, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(18)
```
:::
::: {.callout-note collapse="true" title="Run 019 | Y1 | PF | PERCENT_T1"}
```{r p019, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(19)
```
:::
::: {.callout-note collapse="true" title="Run 020 | Y1 | PF | PERCENT_T2"}
```{r p020, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(20)
```
:::
::: {.callout-note collapse="true" title="Run 021 | Y1 | PF | BLUE_PERCENT"}
```{r p021, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(21)
```
:::
::: {.callout-note collapse="true" title="Run 022 | Y1 | PF | COMP_PERCENT"}
```{r p022, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(22)
```
:::
::: {.callout-note collapse="true" title="Run 023 | Y1 | PF | PC1"}
```{r p023, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(23)
```
:::
::: {.callout-note collapse="true" title="Run 024 | Y1 | PF | PC2"}
```{r p024, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(24)
```
:::

---

## Block D — <span class="badge badge-y1">Y1</span> <span class="badge badge-ffpf">FFPF</span> No Covariate / PF Locus Test (Runs 25–32) {#block-d}

<div class="stat-box">**Purpose:** PF locus appears freely. Compare with Block A to distinguish PF-driven vs PF-independent signals.</div>

::: {.callout-note collapse="true" title="Run 025 | Y1 | FFPF | BIN_T1 | No covariate"}
```{r p025, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(25)
```
:::
::: {.callout-note collapse="true" title="Run 026 | Y1 | FFPF | BIN_T2 | No covariate"}
```{r p026, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(26)
```
:::
::: {.callout-note collapse="true" title="Run 027 | Y1 | FFPF | PERCENT_T1 | No covariate"}
```{r p027, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(27)
```
:::
::: {.callout-note collapse="true" title="Run 028 | Y1 | FFPF | PERCENT_T2 | No covariate"}
```{r p028, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(28)
```
:::
::: {.callout-note collapse="true" title="Run 029 | Y1 | FFPF | BLUE_PERCENT | No covariate"}
```{r p029, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(29)
```
:::
::: {.callout-note collapse="true" title="Run 030 | Y1 | FFPF | COMP_PERCENT | No covariate"}
```{r p030, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(30)
```
:::
::: {.callout-note collapse="true" title="Run 031 | Y1 | FFPF | PC1 | No covariate"}
```{r p031, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(31)
```
:::
::: {.callout-note collapse="true" title="Run 032 | Y1 | FFPF | PC2 | No covariate"}
```{r p032, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(32)
```
:::

---

<div class="banner-comb">📅 Combined Year Results (Y1 + Y2, Year-Adjusted BLUEs)</div>

## Block E — <span class="badge badge-comb">Combined</span> <span class="badge badge-ffpf">FFPF</span> + PF Covariate (Runs 33–40) {#block-e}

<div class="stat-box">**Purpose:** Most powerful block for year-stable PF-independent QTL. Year-adjusted BLUEs from `PERCENT ~ Year + Block + Timepoint + NAME`. QTL here that also appear in Block A are the highest-priority candidates.</div>

::: {.callout-note collapse="true" title="Run 033 | Combined | FFPF | BIN_T1 | Cov: PF"}
```{r p033, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(33)
```
:::
::: {.callout-note collapse="true" title="Run 034 | Combined | FFPF | BIN_T2 | Cov: PF"}
```{r p034, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(34)
```
:::
::: {.callout-note collapse="true" title="Run 035 | Combined | FFPF | PERCENT_T1 | Cov: PF"}
```{r p035, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(35)
```
:::
::: {.callout-note collapse="true" title="Run 036 | Combined | FFPF | PERCENT_T2 | Cov: PF"}
```{r p036, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(36)
```
:::
::: {.callout-note collapse="true" title="Run 037 | Combined | FFPF | BLUE_PERCENT (year-adjusted) | Cov: PF"}
```{r p037, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(37)
```
:::
::: {.callout-note collapse="true" title="Run 038 | Combined | FFPF | COMP_PERCENT | Cov: PF"}
```{r p038, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(38)
```
:::
::: {.callout-note collapse="true" title="Run 039 | Combined | FFPF | PC1 | Cov: PF"}
```{r p039, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(39)
```
:::
::: {.callout-note collapse="true" title="Run 040 | Combined | FFPF | PC2 | Cov: PF"}
```{r p040, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(40)
```
:::

---

## Block F — <span class="badge badge-comb">Combined</span> <span class="badge badge-ff">FF Only</span> (Runs 41–48) {#block-f}

<div class="stat-box">**Purpose:** Highest-priority block. Year-adjusted BLUEs, FF cohort only — both year effects and PF locus removed. A QTL here that also appears in Block B is the strongest candidate for a genuine chilling-requirement gene.</div>

::: {.callout-note collapse="true" title="Run 041 | Combined | FF | BIN_T1"}
```{r p041, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(41)
```
:::
::: {.callout-note collapse="true" title="Run 042 | Combined | FF | BIN_T2"}
```{r p042, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(42)
```
:::
::: {.callout-note collapse="true" title="Run 043 | Combined | FF | PERCENT_T1"}
```{r p043, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(43)
```
:::
::: {.callout-note collapse="true" title="Run 044 | Combined | FF | PERCENT_T2"}
```{r p044, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(44)
```
:::
::: {.callout-note collapse="true" title="Run 045 | Combined | FF | BLUE_PERCENT (year-adjusted)"}
```{r p045, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(45)
```
:::
::: {.callout-note collapse="true" title="Run 046 | Combined | FF | COMP_PERCENT"}
```{r p046, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(46)
```
:::
::: {.callout-note collapse="true" title="Run 047 | Combined | FF | PC1"}
```{r p047, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(47)
```
:::
::: {.callout-note collapse="true" title="Run 048 | Combined | FF | PC2"}
```{r p048, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(48)
```
:::

---

<div class="banner-rp">📊 REP_PROP Results — Proportion of Reps Showing Budbreak (Runs 49–60)</div>

# REP_PROP Results (Runs 49–60)

<div class="info-box">REP_PROP = fraction of replicate plants showing any budbreak (0.25 = 1 of 4 reps, 1.0 = all reps). More evenly distributed than hard binary; reduces single-rep outlier influence.</div>

## Block A_RP — <span class="badge badge-y1">Y1</span> <span class="badge badge-ffpf">FFPF</span> + PF Covariate (Runs 49–50) {#block-a-rp}

::: {.callout-note collapse="true" title="Run 049 | Y1 | FFPF | REP_PROP_T1 | Cov: PF"}
```{r p049, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(49)
```
:::
::: {.callout-note collapse="true" title="Run 050 | Y1 | FFPF | REP_PROP_T2 | Cov: PF"}
```{r p050, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(50)
```
:::

---

## Block B_RP — <span class="badge badge-y1">Y1</span> <span class="badge badge-ff">FF Only</span> (Runs 51–52) {#block-b-rp}

::: {.callout-note collapse="true" title="Run 051 | Y1 | FF | REP_PROP_T1"}
```{r p051, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(51)
```
:::
::: {.callout-note collapse="true" title="Run 052 | Y1 | FF | REP_PROP_T2"}
```{r p052, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(52)
```
:::

---

## Block C_RP — <span class="badge badge-y1">Y1</span> <span class="badge badge-pf">PF Only</span> (Runs 53–54) {#block-c-rp}

::: {.callout-note collapse="true" title="Run 053 | Y1 | PF | REP_PROP_T1"}
```{r p053, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(53)
```
:::
::: {.callout-note collapse="true" title="Run 054 | Y1 | PF | REP_PROP_T2"}
```{r p054, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(54)
```
:::

---

## Block D_RP — <span class="badge badge-y1">Y1</span> <span class="badge badge-ffpf">FFPF</span> No Covariate (Runs 55–56) {#block-d-rp}

::: {.callout-note collapse="true" title="Run 055 | Y1 | FFPF | REP_PROP_T1 | No covariate"}
```{r p055, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(55)
```
:::
::: {.callout-note collapse="true" title="Run 056 | Y1 | FFPF | REP_PROP_T2 | No covariate"}
```{r p056, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(56)
```
:::

---

## Block E_RP — <span class="badge badge-comb">Combined</span> <span class="badge badge-ffpf">FFPF</span> + PF Covariate (Runs 57–58) {#block-e-rp}

::: {.callout-note collapse="true" title="Run 057 | Combined | FFPF | REP_PROP_T1 | Cov: PF"}
```{r p057, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(57)
```
:::
::: {.callout-note collapse="true" title="Run 058 | Combined | FFPF | REP_PROP_T2 | Cov: PF"}
```{r p058, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(58)
```
:::

---

## Block F_RP — <span class="badge badge-comb">Combined</span> <span class="badge badge-ff">FF Only</span> (Runs 59–60) {#block-f-rp}

::: {.callout-note collapse="true" title="Run 059 | Combined | FF | REP_PROP_T1"}
```{r p059, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(59)
```
:::
::: {.callout-note collapse="true" title="Run 060 | Combined | FF | REP_PROP_T2"}
```{r p060, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(60)
```
:::

---

<div class="banner-ord">📊 ORD Results — Ordered Categorical Budbreak (Runs 61–72)</div>

# ORD Results (Runs 61–72)

<div class="info-box">**ORD_T1/T2:** 0 = no reps broke bud, 1 = 1–2 reps broke bud, 2 = 3–4 reps broke bud. Captures the threshold between marginal and consistent budbreak. Added per advisor suggestion to explore 0 / 1–2 / 3+ phenotype binning.</div>

## Block A_ORD — <span class="badge badge-y1">Y1</span> <span class="badge badge-ffpf">FFPF</span> + PF Covariate (Runs 61–62) {#block-a-ord}

::: {.callout-note collapse="true" title="Run 061 | Y1 | FFPF | ORD_T1 | Cov: PF"}
```{r p061, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(61)
```
:::
::: {.callout-note collapse="true" title="Run 062 | Y1 | FFPF | ORD_T2 | Cov: PF"}
```{r p062, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(62)
```
:::

---

## Block B_ORD — <span class="badge badge-y1">Y1</span> <span class="badge badge-ff">FF Only</span> (Runs 63–64) {#block-b-ord}

::: {.callout-note collapse="true" title="Run 063 | Y1 | FF | ORD_T1"}
```{r p063, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(63)
```
:::
::: {.callout-note collapse="true" title="Run 064 | Y1 | FF | ORD_T2"}
```{r p064, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(64)
```
:::

---

## Block C_ORD — <span class="badge badge-y1">Y1</span> <span class="badge badge-pf">PF Only</span> (Runs 65–66) {#block-c-ord}

::: {.callout-note collapse="true" title="Run 065 | Y1 | PF | ORD_T1"}
```{r p065, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(65)
```
:::
::: {.callout-note collapse="true" title="Run 066 | Y1 | PF | ORD_T2"}
```{r p066, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(66)
```
:::

---

## Block D_ORD — <span class="badge badge-y1">Y1</span> <span class="badge badge-ffpf">FFPF</span> No Covariate (Runs 67–68) {#block-d-ord}

::: {.callout-note collapse="true" title="Run 067 | Y1 | FFPF | ORD_T1 | No covariate"}
```{r p067, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(67)
```
:::
::: {.callout-note collapse="true" title="Run 068 | Y1 | FFPF | ORD_T2 | No covariate"}
```{r p068, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(68)
```
:::

---

## Block E_ORD — <span class="badge badge-comb">Combined</span> <span class="badge badge-ffpf">FFPF</span> + PF Covariate (Runs 69–70) {#block-e-ord}

::: {.callout-note collapse="true" title="Run 069 | Combined | FFPF | ORD_T1 | Cov: PF"}
```{r p069, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(69)
```
:::
::: {.callout-note collapse="true" title="Run 070 | Combined | FFPF | ORD_T2 | Cov: PF"}
```{r p070, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(70)
```
:::

---

## Block F_ORD — <span class="badge badge-comb">Combined</span> <span class="badge badge-ff">FF Only</span> (Runs 71–72) {#block-f-ord}

::: {.callout-note collapse="true" title="Run 071 | Combined | FF | ORD_T1"}
```{r p071, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(71)
```
:::
::: {.callout-note collapse="true" title="Run 072 | Combined | FF | ORD_T2"}
```{r p072, echo=FALSE, out.width="100%", results="asis"}
show_manhattan(72)
```
:::

---

<div class="banner">🧬 Candidate Gene Analysis — Ra07 QTL Peak</div>

# Candidate Gene Analysis

## Run 001 Context

<div class="stat-box">
**Source:** Significant markers on Ra07 from Run 001 (Block A, Y1, BIN_T1, PF covariate, N = 214).
**Peak region:** Ra07 ~34.4–36.1 Mb. All significant markers follow an additive model.
Signals here are PF-independent. Note: long LD in blackberry means the causal variant
may be anywhere in the region — peak marker location within a gene is not confirmation
of causality. Peak position may shift as Y2 data are incorporated in combined-year runs.
</div>

## Ra07 Peak — Annotated Candidate Genes

```{=html}
<style>
.gene-card { background:#fff; border:1px solid #cde0d4; border-left:5px solid #2e7d52; border-radius:0 8px 8px 0; padding:1em 1.3em; margin-bottom:1em; }
.gene-card.top-hit { border-left-color:#1a5fa8; background:#f4f9ff; }
.gene-card.high-priority { border-left-color:#6d3a9c; background:#faf8ff; }
.gene-card-header { display:flex; align-items:baseline; flex-wrap:wrap; gap:0.5em; margin-bottom:0.4em; }
.gene-id   { font-family:monospace; font-size:0.82em; color:#5f6b7a; background:#f0f5f2; padding:1px 6px; border-radius:4px; }
.gene-name { font-weight:700; font-size:1.05em; color:#1a3a4a; }
.gene-pos  { font-size:0.82em; color:#7f8c8d; margin-left:auto; }
.gene-mech { font-size:0.92em; color:#2c3e50; line-height:1.65; margin-top:0.35em; }
.marker-pill { display:inline-block; font-size:0.75em; font-weight:700; padding:2px 9px; border-radius:10px; margin-right:4px; vertical-align:middle; }
.pill-top  { background:#dbeafe; color:#1e40af; border:1px solid #93c5fd; }
.pill-high { background:#ede9fe; color:#5b21b6; border:1px solid #c4b5fd; }
.pill-std  { background:#d1fae5; color:#065f46; border:1px solid #6ee7b7; }
</style>

<p style="font-size:0.88em;color:#5f6b7a;margin-bottom:1.2em;">
Genes in chromosomal order (Ra07). Border:
<span style="display:inline-block;width:10px;height:10px;background:#1a5fa8;border-radius:2px;vertical-align:middle;margin:0 3px;"></span>Top marker &nbsp;
<span style="display:inline-block;width:10px;height:10px;background:#6d3a9c;border-radius:2px;vertical-align:middle;margin:0 3px;"></span>High priority &nbsp;
<span style="display:inline-block;width:10px;height:10px;background:#2e7d52;border-radius:2px;vertical-align:middle;margin:0 3px;"></span>Supporting candidate
</p>

<div class="gene-card">
  <div class="gene-card-header"><span class="marker-pill pill-std">5 markers</span><span class="gene-id">Ra_g35125.t1</span><span class="gene-name">HDG1 — HD-ZIP IV homeobox TF</span><span class="gene-pos">34,419,631–34,423,773 bp</span></div>
  <div class="gene-mech">Five markers cluster at the base of the Ra07 peak. HD-ZIP IV TFs regulate meristematic cell identity, epidermal fate, and organ initiation. Allelic variation could alter bud meristem competence to respond to dormancy-release signals.</div>
</div>

<div class="gene-card">
  <div class="gene-card-header"><span class="marker-pill pill-std">1 marker</span><span class="gene-id">Ra_g35171.t1</span><span class="gene-name">Cytokinin hydroxylase (CKX-like)</span><span class="gene-pos">34,629,889–34,632,415 bp</span></div>
  <div class="gene-mech">CKX enzymes irreversibly degrade active cytokinins. Elevated cytokinin promotes earlier budbreak. A low-activity allele would reduce the chill-hour threshold before growth resumes — well-supported in Arabidopsis and Populus dormancy models.</div>
</div>

<div class="gene-card">
  <div class="gene-card-header"><span class="marker-pill pill-std">2 markers</span><span class="gene-id">Ra_g35198.t1</span><span class="gene-name">BAHD acyltransferase DCR</span><span class="gene-pos">34,768,353–34,770,501 bp</span></div>
  <div class="gene-mech">Involved in cutin and bud scale wax biosynthesis. Bud scale cutin restricts movement into the meristem during dormancy. Variation could alter bud scale permeability and ease of dormancy release under low chilling conditions.</div>
</div>

<div class="gene-card">
  <div class="gene-card-header"><span class="marker-pill pill-std">1 marker</span><span class="gene-id">Ra_g35256.t1</span><span class="gene-name">Anthocyanidin 5,3-O-glucosyltransferase</span><span class="gene-pos">35,032,820–35,034,238 bp</span></div>
  <div class="gene-mech">Flavonoid glycosylation enzyme. Anthocyanins accumulate during dormancy; their synthesis may be linked to stress-responsive dormancy regulation.</div>
</div>

<div class="gene-card high-priority">
  <div class="gene-card-header"><span class="marker-pill pill-high">2 markers</span><span class="gene-id">Ra_g35325.t1</span><span class="gene-name">SWEET3 — bidirectional sugar transporter</span><span class="gene-pos">35,400,510–35,401,844 bp</span></div>
  <div class="gene-mech">Two markers fall inside this gene. SWEET proteins facilitate sucrose movement across membranes and are key nodes in phloem loading. Chilling-induced starch-to-sucrose conversion provides both energy and meristematic activation signal during dormancy release. An allele with altered transport kinetics could reduce chilling needed before budbreak is energetically supported.</div>
</div>

<div class="gene-card">
  <div class="gene-card-header"><span class="marker-pill pill-std">2 markers</span><span class="gene-id">Ra_g35383.t1</span><span class="gene-name">Pectinesterase / pectinesterase inhibitor 34</span><span class="gene-pos">35,695,899–35,698,786 bp</span></div>
  <div class="gene-mech">Pectinesterases demethylesterify pectin, altering cell wall rigidity. Cell wall remodeling is a physical prerequisite for budbreak. Variation could affect physical loosening of bud scale tissue.</div>
</div>

<div class="gene-card top-hit">
  <div class="gene-card-header"><span class="marker-pill pill-top">Top marker</span><span class="marker-pill pill-top">-log10(p) = 6.56</span><span class="gene-id">Ra_g35429.t1</span><span class="gene-name">Glucan endo-1,3-beta-glucosidase</span><span class="gene-pos">35,981,988–35,983,708 bp</span></div>
  <div class="gene-mech">The genome-wide top marker falls within this gene body with a second marker also inside the locus. Beta-1,3-glucanases degrade callose at plasmodesmata during endo-dormancy, restoring symplastic connectivity and allowing growth-promoting signals to reach the meristem. Given long LD in blackberry, the causal variant may be nearby rather than in the gene body — this should be presented as a candidate locus rather than a confirmed causal gene. The dosage pattern is consistent: genotypes homozygous for the T allele show the highest early budbreak rates.</div>
</div>

<div class="gene-card">
  <div class="gene-card-header"><span class="marker-pill pill-std">1 marker</span><span class="gene-id">Ra_g35446.t1</span><span class="gene-name">B3 domain TF (RAV/ABI3-like)</span><span class="gene-pos">36,088,025–36,089,609 bp</span></div>
  <div class="gene-mech">ABI3/RAV-like B3 TFs integrate ABA signalling with dormancy maintenance. Variation could lower the ABA threshold at which dormancy is released. Sits immediately upstream of the VRN1-like gene — may form a vernalization × ABA regulatory module.</div>
</div>

<div class="gene-card high-priority">
  <div class="gene-card-header"><span class="marker-pill pill-high">4 markers</span><span class="gene-id">Ra_g35447.t1</span><span class="gene-name">VRN1-like B3 domain transcription factor</span><span class="gene-pos">36,090,128–36,092,040 bp</span></div>
  <div class="gene-mech">Four markers cluster inside this gene — the highest marker density per gene in the entire interval. VRN1 in cereals is a master chilling-accumulation switch activated after sufficient cold exposure. In blackberry a VRN1 homolog would directly regulate the chilling threshold required before dormancy release is permitted. Mateos et al. (2024) identified a RiVRN1.1 ortholog in raspberry as a key dormancy candidate, providing direct Rosaceae comparative genomic support. Alexander's flowering time GWAS peak on Ra07 is ~1 Mb from this region — overlap would be biologically meaningful and further supports this gene as the top priority candidate in the window.</div>
</div>
```

<div class="warn-box">
**Candidate interpretation.** All genes identified by co-location and plausible function only — not functionally validated. LD within the ~1.67 Mb window means the causal variant could lie in any gene or intergenic region. Replication in Blocks B and F (FF cohort, Y1 and combined) is the immediate priority before committing to functional follow-up.
</div>

---
